---
title: "TIDYVERESE UNIVERSE"
author: "R CLASS 101"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
editor_options:
  chunk_output_type: console
---

# Knowing the Rmarkdown environment.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

# dplyr

The dplyr package in R is a powerful tool for data manipulation and transformation.

1.  **`filter()`**

Filters rows based on conditions.

```{r}

# Filter rows where mpg > 20

mtcars_filtered <- mtcars %>%
  filter(mpg > 20 )

```

2.  **`select()`**

Selects columns by name.

```{r}
# Select specific columns

mtcars_selected <- mtcars %>% 
  select(mpg, wt, cyl)

```

3.  **`mutate()`**

Adds new variables (columns) or transforms existing ones.

```{r}
# Add a new column

mtcars_mutated <- mtcars %>% 
  mutate(mpg = mpg / wt)

```

4.  **`summarize() or summarise()`**

Creates summary statistics.

```{r}
# Summarize data

mtcars_summary <- mtcars %>% 
  summarize(mean_mpg = mean(mpg), 
            sd_mpg = sd(mpg),
            mean_displacement = mean(disp))


mtcars %>% 
  group_by(cyl) %>% 
  summarise(mean = mean(mpg),
            sd = sd(mpg), 
            maximum = max(mpg),
            minimum = min(mpg))
```

5.  **`arrange()`**

Sorts rows by the values of one or more columns.

```{r}
# Arrange data by mpg in descending order

mtcars_arranged <- mtcars %>% 
  arrange(desc(mpg))

```

6.  **`group_by()`**

Groups data by one or more variables, often used before summarize function.

```{r}
# Group by number of cylinders

mtcars_grouped <- mtcars %>% 
  group_by(factor(cyl))

```

7.  `rename()`

Renames columns.

```{r}
# Rename columns

mtcars_renamed <- mtcars %>% 
  rename(MilesPerGallon = mpg, Weight = wt)

```

8.  **`transmute()`**

Similar to mutate(), but keeps only the new variables. Combines the select and the mutate functions

```{r}
# Create new variable and keep only the new one

mtcars_transmuted <- mtcars %>% 
  transmute(mpg_per_wt = mpg / wt)

```

9.  **`count()`**

Counts the number of occurrences of unique values.

```{r}
# Count the number of cars in each group of cylinders

mtcars_count <- mtcars %>% 
  count(cyl)

```

10. **`slice()`**

Selects rows by position.

```{r}
# Select the first 5 rows

mtcars_sliced <- mtcars %>% 
  slice(4:10)

```

11. **`sample_n() and sample_frac()`**

Samples rows randomly.

```{r}
# Sample 5 random rows

mtcars_sample_n <- mtcars %>% 
  sample_n(10)

# Sample 10% of rows

mtcars_sample_frac <- mtcars %>% 
  sample_frac(0.1)

```

12. `inner_join()`

Joins two data frames by common columns.

```{r}
# Example data frames

df1 <- tibble(id = c(1, 2, 3), value1 = c("A", "B", "C"))
df2 <- tibble(id = c(1, 2, 4),  value2 = c("X", "Y", "Z"))

# Inner join

joined_df <- df1 %>% 
  inner_join(df2, by = "id")

```

13 **`right_join()`** and **`left_join()`**

Includes all rows from the right data frame and matched rows from the left.

```{r}
# Right join

right_join(df1, df2, by = "id")

# Left join

left_join(df1, df2, by = "id")

# full join

full_join(df1, df2, by = "id")
```

`if_else()` and `case_when()`

```{r}
# Re-coding the levels of a variable using the if_else and case_when functions

mtcars |> 
  select(vs, am, gear) |> 
  mutate(vs = if_else(vs == 0, "V-shaped", "Straight"))


# This function allows you to vectorise multiple if_else() statements.

mtcars |> 
  select(vs, am, gear) |> 
  mutate(gear = case_when(gear == 1 ~ "G1", 
                          gear == 2 ~ "G2", 
                          gear == 3 ~ "G3", 
                          gear == 4 ~ "G4", 
                          gear == 5 ~ "G5"))

# You can also mutate more than one variable  in the dataset

mtcars |> 
  select(vs, am, gear) |> 
  mutate(gear = case_when(gear == 1 ~ "G-1", 
                          gear == 2 ~ "G-2", 
                          gear == 3 ~ "G-3", 
                          gear == 4 ~ "G-4", 
                          gear == 5 ~ "G-5"), 
         
         vs = case_when(vs == 0 ~ "V-shaped", 
                        vs == 1 ~ "Straight"), 
         
         am = ifelse(am == 1, "manual", "automatic")
         )

```

`relocate()`

Move columns to new position.

```{r}

mtcars %>% 
  relocate(cyl, .before = last_col())

mtcars %>% 
  relocate(cyl, .before = carb)

mtcars %>% 
  relocate(cyl, .after = qsec)

mtcars %>% 
  relocate(cyl, .after = wt)
```

`rownames_to_column()`

Tidy data does not use rownames, which store a variable outside of the columns. To work with the rownames, first move them into a column.

```{r}

m = mtcars |> rownames_to_column(var = "type_of_car")

mtcars |> has_rownames()

removed_mtcars <- mtcars %>% 
  remove_rownames()

m |> column_to_rownames(var = "type_of_car")


mtcars |> 
  scale() |> 
  dist() |> 
  hclust() |> 
  factoextra::fviz_dend(k = 3, palette = c("red", "yellow", "green"), 
            lwd = 1, horiz = T)
```

# ggplot2

The ggplot2 package in R is very powerful and versatile, offering a wide range of functions for creating complex and customizable visualizations.

1.  **geom\_ Functions**

These functions add different types of layers to a plot:

**`geom_line()`**: Adds a line plot.

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_line()

```

**`geom_bar()`**: Adds a bar plot (for counts).

```{r}
ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar()

```

**`geom_histogram()`**: Adds a histogram.

```{r}
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 2)
```

**`geom_boxplot()`**: Adds a boxplot.

```{r}
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot()
```

**`geom_density()`**: Adds a density plot.

```{r}
ggplot(mtcars, aes(x = mpg)) +
  geom_density()

```

**`geom_smooth()`**: Adds a smoothed conditional mean (e.g., a trend line).

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm")

```

**`geom_violin()`**`:` Adds a violin plot.

```{r}
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_violin()

```

2.  **facet\_ Functions**

These functions create multiple panels in a plot:

**`facet_wrap()`**: Creates a series of plots wrapped into a grid.

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  facet_wrap(~ cyl)
```

**`facet_grid()`**: Creates a grid of plots based on two variables.

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  facet_grid(vs ~ am)

```

3.  **scale\_ Functions**

These functions customize the scales of a plot:

**`scale_x_continuous(), scale_y_continuous()`**: Customize continuous scales.

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  scale_x_continuous(name = "Weight (1000 lbs)") +
  scale_y_continuous(name = "Miles per Gallon")

```

**`scale_fill_manual(), scale_color_manual()`**: Customize colors.

```{r}
ggplot(mtcars, aes(x = factor(cyl), fill = factor(vs))) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("0" = "red", "1" = "blue")) + 
  labs(fill = "Automation", caption = "Car that are automatic(0) and manual(1)")

```

4.  **theme\_ Functions**

These functions customize the appearance of a plot:

**`theme()`**: Customize specific theme elements.

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  theme(
    axis.title.x = element_text(color = "blue"),
    axis.title.y = element_text(color = "red")
  )
```

**`theme_minimal(), theme_classic(), theme_bw()`**: Apply pre-built themes.

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  theme_minimal()

```

5.  ***coord\_ Functions***

These functions customize the coordinate system of a plot:

**`coord_flip()`**: Flips the x and y axes.

```{r}
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot() +
  coord_flip()

```

6.  `labs()`

This function customizes labels for the plot.

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  labs(
    title = "Car Weight vs. MPG",
    x = "Weight (1000 lbs)",
    y = "Miles per Gallon"
  )
```

jyg.sjklrjfgk# Forcats

The forcats package in R is specifically designed for working with categorical variables (factors).

Factors and characters are both data types used to represent non-numeric data, but they have different properties and uses:

***Factors:***

-   Represent categorical data (e.g., gender, color, education level)
-   Stored as integers with labels
-   Have a specific order (which can be defined or default)
-   Used in statistical modeling, data analysis, and visualization
-   Can be converted to characters using as.character()

***Characters:***

-   Represent text or string data (e.g., names, addresses, descriptions)
-   Stored as vectors of characters
-   No inherent order or categorization
-   Used in text manipulation, data cleaning, and string analysis
-   Can be used in concatenation, substring extraction, and regular expressions
-   Can be converted to factors using as.factor()

***Key differences:***

-   Purpose: Factors are used for categorical data, while characters are used for text data.
-   Storage: Factors are stored as integers with labels, while characters are stored as vectors of characters.
-   Order: Factors have a specific order, while characters do not.
-   Analysis: Factors are used in statistical analysis and modeling, while characters are used in text analysis and manipulation.

In summary, factors are used for categorical data and are stored with labels, while characters are used for text data and are stored as vectors of characters.

1.  `fct_reorder()`

Reorders factor levels by a numerical variable.

```{r}
# Reorder the factor levels of cyl by the mean of mpg

mtcars <- mtcars %>% 
  mutate(cyl = factor(cyl))

mtcars <- mtcars %>% 
  mutate(cyl = fct_reorder(cyl, mpg))

# Plot to show reordered factor levels

ggplot(mtcars, aes(x = cyl, y = mpg)) +
  geom_boxplot()


```

2.  `fct_lump()`

Lumps together the least common levels into a single "Other" level.

```{r}
# Lump together the least common levels into 'Other'

mtcars_lump <- mtcars %>% 
  mutate(cyl = fct_lump(factor(cyl), n = 1)) %>% 
  count(cyl)

# Check the levels after lumping

levels(mtcars_lump$cyl)



```

3.  `fct_collapse()`

Collapses multiple levels into a single level.

```{r}
# Collapse multiple levels into a single level

mtcars_collapse <- mtcars %>% 
  mutate(cyl = fct_collapse(factor(cyl), "Large" = c("6", "8")))

# Check the levels after collapsing

levels(mtcars_collapse$cyl)

```

# lubridate

The lubridate package in R is designed for working with dates and times, simplifying the process of handling and manipulating temporal data.

1.  `ymd(), ymd_hms()`, etc.

Parse dates and times from strings.

```{r}
# Parse date

date <- ymd("2021/01/01")

# Parse date and time

datetime <- ymd_hms("2021-01-01 12:34:56")

```

2.  `now()`

Get the current date and time.

```{r}
# Current date and time

current_time <- now()

```

3.  `today()`

Get the current date.

```{r}
# Current date

current_date <- today()

```

4.  `year(), month(), day(), hour(), minute(), second()`

Extract components from a date-time object.

```{r}
# Extract year, month, day, hour, minute, second

year_value <- year(datetime)

month_value <- month(datetime)

day_value <- day(datetime)

hour_value <- hour(datetime)

minute_value <- minute(datetime)

second_value <- second(datetime)

```

# stringi

The stringi package in R is designed for consistent, simple, and convenient string manipulation.

1.  `str_length()`

Returns the number of characters in a string.

```{r}
# Get the length of each string

lengths <- str_length(c("apple", "banana", "cherry"))

```

2.  `str_c()`

Concatenates strings.

```{r}
# Concatenate strings

concatenated <- str_c("Hello", " ", "World", "!")

```

3.  `str_detect()`

Detects the presence of a pattern in strings.

```{r}
# Detect pattern

has_pattern <- str_detect(c("apple", "banana", "cherry"), "a")

```

4.  `str_count()`

Counts the number of occurrences of a pattern in strings.

```{r}
# Count occurrences of a pattern

count <- str_count(c("apple", "banana", "cherry"), "a")

```

5.  `str_to_upper()` and `str_to_lower()`

Converts strings to upper or lower case.

```{r}
# Convert to upper case

upper <- str_to_upper("hello")

# Convert to lower case

lower <- str_to_lower("HELLO")




```

6.  `str_to_title()`

Converts strings to title case.

```{r}
# Convert to title case

title <- str_to_title("the quick brown fox")

```

# readr

The readr package in R is designed for fast and friendly reading of rectangular data, such as CSV files into and out of the R environment.

# tibble

The tibble package in R is designed to provide a modern take on data frames, making data manipulation easier and more intuitive.

# tidyr

The tidyr package in R is designed to help tidy your data, making it easier to work with.

1.  `pivot_longer()`

Converts data from wide to long format.

```{r}
# Convert data from wide to long format

long_data <- pivot_longer(mtcars, cols = c(mpg, disp), names_to = "variable", values_to = "value")

```

2.  `pivot_wider()`

Converts data from long to wide format.

```{r}
# Convert data from long to wide format

wide_data <- pivot_wider(long_data, names_from = variable,
                         values_from = value)
```

3.  `drop_na()`

Drops rows containing missing values.

```{r}
library(missMDA)
data(snorena)

# Drop rows with missing values

clean_data <- drop_na(snorena)
```

Additional Information

```{r}
library(equatiomatic)

# Fit a simple model
mod1 <- lm(mpg ~ factor(cyl) + disp, mtcars)

# Give the results to extract_eq
extract_eq(mod1) 
```

```{r}
extract_eq(mod1, use_coefs = T, intercept = "gamma", swap_var_names = c("cyl" = "Cylinders", "disp" = "Displacement"))
```

$$
\operatorname{mpg} = \alpha + \beta_{1}(\operatorname{factor(cyl)}_{\operatorname{6}}) + \beta_{2}(\operatorname{factor(cyl)}_{\operatorname{8}}) + \beta_{3}(\operatorname{disp}) + \epsilon
$$

$$
\operatorname{\widehat{mpg}} = 34.66 - 1.59(\operatorname{Cylinders}) - 0.02(\operatorname{Displacement})
$$
